---
title: "Quantifying Parameter Uncertainty with ipmr"
output:
  rmarkdown::html_vignette: 
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Quantifying Parameter Uncertainty with ipmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

Integral projection models (IPMs) are commonly used to estimate population growth rates and quantify the demographic importance of different vital rates. However, parameter estimates used to construct IPMs are themselves uncertain, and this uncertainty propagates into model predictions. The `sensitivity()` function estimates how population growth rate (λ) responds to changes in individual model parameters, while `uncertainty()` uses these sensitivities together with the covariance structure of sampled parameter estimates to estimate how parameter uncertainty contributes to uncertainty in λ.  

This vignette demonstrates how to:  
- perform a sensitivity analysis
- prepare sampled parameter values  
- perform an uncertainty analysis  
- interpret the output  
- visualize parameter and vital rate contributions to uncertainty 

## Sensitivity analysis {#sensitivity}

The `sensitivity()` function calculates the sensitivity and elasticity of λ to changes in individual model parameters using finite differences. Sensitivity describes the absolute change in λ associated with a change in a parameter, whereas elasticity expresses this relationship on a proportional scale. 

### Performing the analysis with `sensitivity()`

A sensitivity analysis requires an `ipmr` model object and a specification of the IPM matrix through the `mega_mat` argument. Parameters can optionally be specified using the pars argument.

```{r, eval = F, echo = T}
sens <- sensitivity(
  ipm = ipm,
  pars = pars,
  mega_mat = mega_mat
)
```

If `pars = NULL`, sensitivities are calculated for all parameters in the IPM.

#### The `mega_mat` argument {#mega-mat}

The `mega_mat` argument specifies how the subkernels of the IPM should be assembled into the iteration matrix used to calculate λ. It is passed directly to `make_iter_kernel()`, which constructs the complete iteration matrix from the subkernels stores in the `ipmr` object.

For **simple IPMs**, the iteration kernel is constructed automatically by summing the model's subkernels. In this case, `mega_mat` does not need to specified explicitly.

For **general IPMS**, `mega_mat` specifies the complete block structure of the iteration matrix. Matrix blocks must be supplied in **row-major order**, meaning that entries are specified from left to right across the first row, followed by the second row, and so on.

For example, a two-state-class IPM with the block structure

[
\mathbf{K} =
\begin{pmatrix}
\mathbf{P}{11} & \mathbf{F}{12} \
\mathbf{F}{21} & \mathbf{P}{22}
\end{pmatrix}
]

would be specified as:

```{r, eval = F, echo = T}
mega_mat <- c(
  "P_11", "F_12",
  "F_21", "P_22"
)
```

The four entries correspond to the four blocks of the matrix in row-major order:

`P_11   F_12
F_21   P_22`

The entries in `mega_mat` must correspond to subkernels that exist in the ipmr object. The names should therefore match the kernel names used when constructing the IPM.

##### Combining multiple subkernels within a block

A single matrix block can contain the sum of multiple subkernels. This is useful when more than one demographic process contributes to the same transition between state classes.

For example, suppose the upper-left block receives contributions from both `P` and `F`, while the lower-right block contains `P_22`. The iteration matrix could be written as

[
\mathbf{K} =
\begin{pmatrix}
\mathbf{P} + \mathbf{F} & \mathbf{B} \
\mathbf{C} & \mathbf{P}_{22}
\end{pmatrix}.
]

The corresponding mega_mat specification is:

```{r, eval = F, echo = T}
mega_mat <- c(
  "P + F", "B",
  "C",     "P_22"
)
```

Alternatively, the entire specification can be supplied as a character string:

`mega_mat <- "c(P + F, B, C, P_22)"`

Both forms use the same syntax internally. The string form can be convenient when the matrix specification is generated programmatically or stored as a single object.

In most cases, users do not need to construct these matrices manually. The important requirement is that the `mega_mat` specification accurately represents the biological transitions encoded by the IPM and uses the same kernel names as the original model.

##### Other specifications accepted by `make_iter_kernel()`

As the `mega_mat` argument is passed only to `make_iter_kernel()`, other specifications are accepted as described in the documentation for \code{\link{make_iter_kernel}}. These include parameter set indices, the identity matrix `I`, and the zero matrix `0`. 

#### Numerical differentiation

Sensitivities are calculated using finite differences. By default, `sensitivity()` perturbs each parameter by `delta = 1e-4` and uses a central finite difference:

[
\frac{\lambda(\theta + \delta) - \lambda(\theta - \delta)}
{2\delta}.
]

The perturbation size should be small enough to provide a useful numerical approximation while remaining large enough to avoid numerical precision problems.

For parameters with biological bounds, such as probabilities constrained to the interval [0, 1], subtracting `delta` may produce an invalid parameter value or kernel construction. The bounds argument can be used to specify these constraints:

```{r, eval = F, echo = T}
sens <- sensitivity(
  ipm = ipm,
  mega_mat = c("P", "F"),
  bounds = list(
    s_int    = c(-Inf, Inf), #default behavior; does not need to be specified
    survival = c(0, 1)
  )
)
```

Bounds are specified as a named list, with each element containing the lower and upper bound for one parameter. Parameters not included in `bounds` are treated as unbounded.

When a parameter is sufficiently close to a specified boundary that a central perturbation would cross that boundary, `sensitivity()` automatically uses a one-sided finite difference. For example, if a parameter is exactly at its lower bound, a forward difference is used:

[
\frac{\lambda(\theta + \delta) - \lambda(\theta)}
{\delta}.
]

Similarly, a backward difference is used when perturbation toward the upper side would cross an upper bound.

The finite-difference method used for each parameter is returned in the `difference_method` element of the result.

#### Interpreting sensitivity output

The result of `sensitivity()` contains sensitivity and elasticity estimates for each parameter:

`sens$sensitivity
sens$elasticity
sens$difference_method`

Sensitivity values describe the absolute effect of a parameter on λ, whereas elasticity values describe the proportional effect. Elasticity is therefore particularly useful when comparing parameters measured on different scales.

The `difference_method` output indicates whether each parameter was evaluated using a central, forward, or backward finite difference.

Sensitivity analysis can be useful as a standalone analysis for identifying parameters to which population growth is most responsive. It also provides the parameter sensitivities used by `uncertainty()` to decompose uncertainty in λ.

## Quick start guide {#quick}  

To perform an uncertainty analysis with `uncertainty()`, you will need four required inputs and four optional inputs.  

### `ipm`  

An `ipmr` model object returned by `make_ipm()`. Currently, `uncertainty()` supports **simple** and **general**, **d**ensity-**i**ndependent, **det**erministic IPMs. Stochastic and density-dependent IPMs are not currently supported.  

### `pars`  

A character vector giving the names of the parameters to examine in the uncertainty analysis. In most applications, this should include every parameter that varies among the sampled parameter sets.  

If `pars = NULL`, all parameters used by the IPM are included. Parameters that do not vary across samples will simply have zero estimated uncertainty contribution.  

### `samples`  

A dataframe containing sampled parameter values. Each row should represent one complete parameter set (one sampled model), and each column should contain the sampled values of a single model parameter. Column names must exactly match the parameter names used to construct the `ipmr` object.  

Sampled parameter values may be obtained by bootstrap resampling demographic data and refitting the vital rate models, or by drawing from the joint posterior distribution of a Bayesian model. Guidance for preparing these samples is provided in Section @ref(bootstrap).  

All parameters required to construct the IPM must be included in `samples`, even if some parameters are held constant across all sampled models. We recommend using at least 1,000 sampled parameter sets to obtain stable estimates of parameter uncertainty.  

### `mega_mat`  

A character vector or string specifying the subkernel structure of the `ipmr` object. For general IPMs, the `mega_mat` argument describes how subkernels are assembled into the complete iteration matrix. The subkernel names must match those used to construct the `ipmr` object and must be supplied in row-major order.  

The same `mega_mat` specification used for `sensitivity()` should be supplied to `uncertainty()` when both analyses are performed on the same IPM. See the [discussion of `mega_mat`]{#mega-mat} in the sensitivity analysis section for examples.

### `vr_table` (optional)  

An optional dataframe mapping each parameter to the vital rate to which it corresponds. This table is used to summarize parameter uncertainty contributions by vital rate and to support downstream visualization with the `plot()` function. If omitted, uncertainty contributions are only reported at the parameter level.

Calling `uncertainty(ipm, vr_table = "template")` generates a template dataframe that can be manually filled in for all parameters in the IPM. 

### `delta` (optional)  

A numeric scalar specifying the perturbation size used for numerical differentiation when calculating parameter sensitivities. The default is `delta = 1e-4`.

For some model constructions, selecting a `delta` value greater than the absolute value of the smallest parameter value may result in invalid subkernels. Care should be taken to select a sufficiently small value for `delta` such that perturbation of the sampled parameter values remains within the biologically feasible parameter space.

When parameter bounds are known, they can be supplied through the `bounds` argument. `uncertainty()` then uses the same finite-difference rules as `sensitivity()`, including forward or backward differences when a perturbation would cross a specified boundary.

### `bounds` (optional)

An optional named list specifying lower and upper bounds for model parameters. Each element should be a numeric vector of length two giving the lower and upper bounds, respectively. 

For example:
```{r, eval = F, echo = T}
bounds <- list(
  survival    = c(0, 1),
  recruitment = c(0, Inf)
)
```

Parameters omitted from `bounds` are treated as unbounded. When sampled parameter values occur within `delta` of a boundary, `uncertainty()` automatically selects an appropriate one-sided finite difference rather than perturbing the parameter outside its feasible range.  

The supplied bounds must apply to every sampled value of the corresponding parameter. Sampled values outside the specified bounds are rejected because the resulting sensitivity calculations would no longer represent the specified parameter space.

### `cores` (optional)  

The number of processor cores to use during the analysis. The default (`cores = 1`) performs the analysis sequentially. For larger IPMs or many sampled parameter sets, parallel execution can substantially reduce computation time.  

We recommend using

```{r, eval = F, echo = T}
parallelly::availableCores(omit = 1)
```

to determine the number of available cores while leaving one core free for other system processes.  

Once these objects have been prepared, the uncertainty analysis can be performed with a single function call:

```{r, eval = F, echo = T}
unc <- uncertainty(
  ipm = ipm,
  samples = samples,
  pars = pars,
  mega_mat = mega_mat,
  vr_table = vr_table,
  cores = parallelly::availableCores(omit = 1)
)
```

The resulting object contains the sampled population growth rates, estimates of total model uncertainty, and summaries of uncertainty contributions at both the parameter and (optionally) vital rate levels.  

## Understanding the output {#output}  

`uncertainty()` returns a list containing the original IPM, the sampled population growth rates, and summaries of uncertainty at both the parameter and vital rate levels. 

### `ipm`  

The original IPM object supplied to `uncertainty()`. This is returned unchanged for convenience and reproducibility.  

### `lambdas`  

A numeric vector containing the population growth rates (λ) calculated from each of the sampled parameter sets. The variance of this vector represents the estimated overall model uncertainty.  

### `mod_uncert`  

A single numeric value giving the estimated model uncertainty, calculated as the variance of the sampled λ values (`var(lambdas)`).  

### `params_uncert` 

A dataframe summarizing parameter-level uncertainty. Each row corresponds to a single model parameter and includes:  
- the mean sensitivity and elasticity across all sampled models, and the finite-difference method used to calculate the sensitivities
- the variance, standard error, and coefficient of variation of the sampled parameter values
- the estimated contribution of the parameter to the uncertainty in λ
- the proportion of total model uncertainty explained by the parameter
- the vital rate the parameter corresponds to (if provided)

The parameter uncertainty contributions are calculated by combining the parameter covariance matrix with the average sensitivities estimated across the sampled models.  

### `vr_uncert`  

If a `vr_table` is supplied, `vr_uncert` summarizes parameter uncertainty contributions by vital rate. Parameter contributions are summed within each vital rate, allowing users to identify which demographic processes contribute most to uncertainty in the population growth rate. If `vr_table` is omitted, this element contains `NA`.  

## Obtaining sampled parameter values {#bootstrap}  

The `samples` argument should contain a representative sample from the joint distribution of the model parameters. Because `uncertainty()` combines parameter sensitivities with the covariance structure of the sampled parameter values, it is important that correlations among parameters are preserved.  

Sampled parameter values may be obtained using either Bayesian or frequentist methods.  

### Bayesian models  

When vital rate models are fit within a Bayesian framework, sampled parameter values can be obtained directly by drawing from the joint posterior distribution. Each posterior draw represents one complete parameter set and can be supplied directly to `uncertainty()` after arranging the draws into a dataframe whose column names match the parameter names used in the IPM.  

### Bootstrap resampling  

For models fit using maximum likelihood or other frequentist methods, we recommend generating sampled parameter values using bootstrap resampling. Briefly, bootstrap resampling repeatedly samples the original demographic data with replacement, refits each vital rate model to the resampled data, and stores the resulting parameter estimates. Repeating this procedure produces a distribution of parameter values while preserving the covariance structure among parameters estimated from the same resampled dataset.  

Although the exact implementation will depend on the study design and modelling framework, the following recommendations will help produce sampled parameter values that are compatible with `uncertainty()`. 

#### Keep constants fixed across bootstrap samples  

Some quantities used to construct an IPM, such as the lower and upper integration limits (`L` and `U`), are determined by the observed range of the original dataset rather than estimated by a statistical model. These quantities should generally be held constant across bootstrap samples.  

Recalculating such values for every bootstrap replicate may produce artificial variability arising from differences in the sampled observations rather than genuine uncertainty in the demographic process.  

#### Use a consistent model structure  

If model selection is performed before constructing the IPM, it should be performed only once using the original dataset. The selected model structure should then be refit to every bootstrap sample.  

Allowing the model structure to change among bootstrap replicates introduces model selection uncertainty in addition to parameter uncertainty. Because `uncertainty()` is designed to quantify parameter uncertainty, all sampled parameter sets should be generated from the same underlying model structure.  

#### Consider the sampling design   

Bootstrap resampling should reflect the biological and statistical structure of the original study whenever possible. Depending on the study design, it may be appropriate to stratify sampling by factors such as year, site, or experimental block so that each bootstrap replicate preserves important features of the data. However, it is important to consider the tradeoff between stratification structure and the resulting sample size of each resampled stratified unit.  

The appropriate resampling strategy depends on the size and structure of the dataset, and no single approach is appropriate for all applications. Users should select a strategy that is consistent with the assumptions of their demographic models and the biological processes being studied. 

#### Take caution when independently resampling multiple datasets  

Integral projection models frequently combine information from multiple demographic datasets. For example, survival, growth, and reproduction may be estimated from a long-term monitoring study, while fecundity or recruitment may be estimated from a separate, one-off experiment.  

When these datasets are bootstrapped independently, parameter covariances between demographic processes cannot be accurately estimated because each dataset is resampled separately. Consequently, the covariance terms used in the uncertainty analysis may not reflect biologically-relevant covariance between these terms, or the true covariance may be much higher than estimated, and the resulting uncertainty contributions must be interpreted with caution.  

This limitation is not unique to `uncertainty()` but reflects the information available in the sampled parameter values. Whenever possible, uncertainty analyses should therefore be based on parameter samples generated from a single integrated dataset or from a modelling framework that estimates the full joint parameter distribution. 


## Plotting the results {#viz}

The output of `uncertainty()` can be passed directly to `plot()` to visualize the estimated contributions of parameters or vital rates to model uncertainty using `ggplot2`.  

By default, `plot()` returns a list containing all available visualizations. Individual plots may also be specified using the `type` argument. `type = "vr"` and `type = "stacked"` are only available if `vr_table` is specified. 

### Parameter-level uncertainty  

```{r, eval = F, echo = T}
plot(uncert_out, type = "param")
```

The parameter-level plot summarizes the uncertainty analysis for every parameter specified in the analysis. Three panels are displayed:  
- **Parameter uncertainty**, represented as the variance of the sampled parameter values  
- **Model sensitivity**, shown as the mean sensitivity of λ to each parameter across all sampled models
- **Parameter contribution to uncertainty**, showing the estimated proportion of model uncertainty attributable to each parameter  

Parameters are grouped and colored according to the vital rate to which they correspond, when supplied, making it easy to distinguish whether uncertainty arises from variation within a vital rate or differences among vital rates.  

### Vital rate uncertainty  

```{r, eval = F, echo = T}
plot(uncert_out, type = "vr")
```

When a `vr_table` is supplied to `uncertainty()`, parameter uncertainty contributions are summed within each vital rate. This plot provides a higher-level summary by showing the total contribution of each demographic process to uncertainty in λ.  

This visualization is useful for identifying which processes of the life cycle contribute most to uncertainty in model projections.  

### Model uncertainty decomposition  

```{r, eval = F, echo = T}
plot(uncert_out, type = "stacked")
```

The stacked uncertainty plot summarizes the uncertainty contributions of all vital rates within a single bar. Colored segments represent the uncertainty attributed to each vital rate, while the outlined box indicates the total variance in λ estimated from the sampled models.  

This visualization provides a quick assessment of how much of the total model uncertainty is explained by the parameter uncertainty captured in the analysis.  

### Customizing colors  

By default, the plotting functions use a colorblind-friendly palette to distinguish vital rates. Custom colors can be supplied using the `vr_colors` argument as a named character vector, where the names correspond to the values in the `vital_rate` column of the uncertainty output.  


Note: when exporting figures as PDF files using `ggsave()`, use `device = cairo_pdf` to preserve the λ character in the figure text. 
